---
title: "M. bovis Pangenome Analysis"
author: "Noah Legall"
date: "9/22/2020"
output:
  pdf_document: default
  html_document: default
---

## Introduction

Here, I am doing a full-fledged PCA analysis on the Accessory genome of M. bovis to see how isolates cluster together. 

## Figures

# 1. Data Description

```{r}
###1. This is the metadata we pulled from NCBI. Make sure your working directory is the Resource folder on this 
library(ggpubr)
library(dplyr)
accession_meta <- read.csv("ProjectBovisReservoir_metadata.csv", header = TRUE, stringsAsFactors = FALSE)

accession_bar <- accession_meta %>% group_by(Species, Host) %>% summarise(no_in_data = length(Species))

accession_bar <- accession_bar[order(accession_bar$Host,accession_bar$no_in_data),]

#A bar plot separated by species, colored by reservoir status.
ggbarplot(accession_bar,"Species","no_in_data",  fill = "steelblue", color = "steelblue", xlab = "Host Species", ylab = "No. of isolates", lab.size = 10) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),axis.text=element_text(size = 15), axis.title = element_text(size = 18)) +
scale_y_log10()
```

# 2. Assembly Stat Distributions

```{r}
library(patchwork)
library(magrittr)
setwd("/Users/noah_/Documents/Bovis-PangenomeOfReservoirs/Resources/")
assembly_stats <- read.csv("mbovis_transposed_report.csv",header = TRUE)
assembly_stats %<>%
  filter(N50 > 45000) %>%
  filter(Total.length < 5000000) %>%
  filter(GC.... > 65.50 & GC.... < 65.6) %>%
  filter(Genome.fraction.... > 97.0)

N50_hist <- assembly_stats %>% ggplot(aes(x=N50)) +
                   geom_histogram(binwidth = 5000, fill = "steelblue", color = "steelblue") +
                   theme_light()

genome_length_hist <- assembly_stats  %>% ggplot(aes(x=Total.length)) +
  geom_histogram(binwidth = 1000, fill = "steelblue", color = "steelblue") +
  xlab("Genome Length (bp)") +
  theme_light()

GC_hist <- assembly_stats %>% ggplot(aes(x=GC....)) +
  geom_histogram(binwidth = 0.005, fill = "steelblue", color = "steelblue") +
  xlab("GC content") +
  theme_light()

genome_frac_hist <- assembly_stats  %>% ggplot(aes(x=Genome.fraction....)) + 
  geom_histogram(binwidth = 0.01, fill = "steelblue", color = "steelblue") +
  xlab("Genome Fraction") +
  theme_light()

(N50_hist | genome_length_hist)/(GC_hist | genome_frac_hist) 

```

# 3. Accessory Genome Presence/Absence Matrix

```{r}
library(dplyr)
gene_pres_abs <- read.csv("mbovis_prab.csv", header = TRUE, stringsAsFactors = FALSE, row.names = "Gene")
accessory_genome <- gene_pres_abs[!(is.na(gene_pres_abs$Accessory.Fragment)),]
core_genome <- gene_pres_abs[is.na(gene_pres_abs$Accessory.Fragment),]

accessory_genome_non_unique <- accessory_genome
auxil <- gene_pres_abs %>% select(2:14)

accessory_pa <- accessory_genome_non_unique %>% select(14:(ncol(accessory_genome_non_unique)))
##edit the gene presence absence to be numeric.
accessory_pa[!(accessory_pa=="")] <- 1
accessory_pa[accessory_pa==""] <- 0
```

```{r}
##let's transpose the dataframe by turning it into a matrix first. 
pa_transpose <- t(data.matrix(accessory_pa))
heatmap(pa_transpose, scale = "none", Rowv = NA, Colv = NA, col = c("white","blue"), main = "Mycobacterium bovis Accessory Genome",labRow = NULL,labCol = NULL)

```


# 4. Principal Component Analysis 
```{r}
# for this we just need to convert the full PRAB matrix into a PCA plot. I want to know about the clustering, and what variables contribute to that clustering. - per the suggestion of Dr. Liu
library("ggplot2")
prab_pca <- prcomp(pa_transpose)
variance <- (prab_pca$sdev)^2
loadings <- prab_pca$rotation
rownames(loadings) <- colnames(pa_transpose)
scores <- prab_pca$x 
```

```{r}
varPercent <- variance/sum(variance) * 100 
prin_comp <- paste0("PC",1:10)
data <- data.frame(prin_comp,varPercent[1:10])
ggplot(data, aes(x=reorder(prin_comp, -varPercent[1:10]), y=varPercent[1:10], )) +
  geom_bar(stat = "identity",fill='steelblue', color="steelblue") +
  geom_text(aes(label=paste0(signif(varPercent[1:10],3)," %")), vjust=1.6, color="white", size=3.5) +
  labs(x = "Principal Components", y= "% of Explained Variance") +
  ggtitle("PCA - Scree Plot") +
  theme_minimal()
```

```{r}
# I am curious how the clustering looks for the top 4 PC's. For the PCoA, we notice some strange clustering in the first 2 components, so I want to investigate if that is true as well. 

mbov_full_meta <- read.csv("../../../Documents/Mbovis_meta.csv", header = TRUE,stringsAsFactors = FALSE)
mbov_meta <- read.csv("/Users/noah_/Documents/filtered_isolate_list.csv",header = TRUE, stringsAsFactors = FALSE)
upd_mbov_meta <- mbov_meta %>% left_join(mbov_full_meta %>% select(Experiment,Instrument,Center.Name,Collection_Date), by = c("Sample" = "Experiment"))
mbov_meta <- upd_mbov_meta

scores <- as.data.frame(scores[,1:5])
scores$country <- mbov_meta$Country
scores$species <- mbov_meta$Species
scores$instrument <- mbov_meta$Instrument

country.clust <- ggplot(as.data.frame(scores),aes(x=PC3,y=PC4, fill = country, col = country)) +
  geom_point() +
  theme_minimal()

species.clust <- ggplot(as.data.frame(scores),aes(x=PC3,y=PC4, fill = species, col = species)) +
  geom_point() +
  theme_minimal()

species.clust

#looks just like the PCoA! so I think I will just work with this hence forth.
```

```{r}
#Let's explore the 1st two PC's to see what can be done to resolve the clustering confusion.

# Let's relook at PC1 + PC2
ggplot(as.data.frame(scores),aes(x=PC1,y=PC2, fill = species, col = species)) +
  geom_point() +
  theme_minimal()
```



```{r}
ggplot(as.data.frame(scores),aes(x=PC1,y=PC3, fill = instrument, col = instrument)) +
  geom_point() +
  theme_minimal()

ggplot(as.data.frame(scores),aes(x=PC2,y=PC3, fill = country, col = country)) +
  geom_point() +
  theme_minimal()
```
It seems that PC2 is the strange actor in this analysis. I observed this during the summer and just opted to remove PC's 1 & 2. Without knowing too much however, I think it might be harder to explain removing both PC's so I will argue for keeping PC 1, and then analyzing PC3. However, the clustering seems to be amazing for PC 3 & 4. So what should be done? I will use Variable loadings with the PC's so this might be an important decision (but not one I want to belabor the point over)

```{r}
ggplot(as.data.frame(scores),aes(x=PC3,y=PC4, fill = species, col = species)) +
  geom_point() +
  theme_minimal()
```

Next I wonder how the clustering looks given different thresholds of accessory genome presence. What part of the accessory genome explains the clustering that we see?

```{r}
soft_core_genes <- subset(accessory_genome,No..isolates > 665 & No..isolates < 693) %>% select(14:(ncol(accessory_genome)))
soft_core_genes[!(soft_core_genes=="")] <- 1
soft_core_genes[soft_core_genes==""] <- 0
soft_core_prab <- t(data.matrix(soft_core_genes))

shell_genes <- subset(accessory_genome,No..isolates > 105 & No..isolates < 665) %>% select(14:(ncol(accessory_genome)))
shell_genes[!(shell_genes=="")] <- 1
shell_genes[shell_genes==""] <- 0
shell_prab <- t(data.matrix(shell_genes))

cloud_genes <- subset(accessory_genome,No..isolates > 0 & No..isolates < 105) %>% select(14:(ncol(accessory_genome)))
cloud_genes[!(cloud_genes=="")] <- 1
cloud_genes[cloud_genes==""] <- 0
cloud_prab <- t(data.matrix(shell_genes))
```

Now we will just redo the analysis we did with all the data:

```{r}
prab_pca <- prcomp(soft_core_prab)
variance <- (prab_pca$sdev)^2
loadings <- prab_pca$rotation
rownames(loadings) <- colnames(soft_core_prab)
scores <- prab_pca$x 

varPercent <- variance/sum(variance) * 100 
prin_comp <- paste0("PC",1:10)
data <- data.frame(prin_comp,varPercent[1:10])
ggplot(data, aes(x=reorder(prin_comp, -varPercent[1:10]), y=varPercent[1:10], )) +
  geom_bar(stat = "identity",fill='steelblue', color="steelblue") + 
  labs(x = "Principal Components", y= "% of Explained Variance") +
  ggtitle("PCA - Scree Plot") +
  theme_minimal()

scores <- as.data.frame(scores[,1:4])
scores$country <- mbov_meta$Country
scores$species <- mbov_meta$Species

ggplot(as.data.frame(scores),aes(x=PC3,y=PC4, fill = country, col = country)) +
  geom_point() +
  theme_minimal()

prab_pca <- prcomp(shell_prab)
variance <- (prab_pca$sdev)^2
loadings <- prab_pca$rotation
rownames(loadings) <- colnames(shell_prab)
scores <- prab_pca$x 

varPercent <- variance/sum(variance) * 100 
prin_comp <- paste0("PC",1:10)
data <- data.frame(prin_comp,varPercent[1:10])
ggplot(data, aes(x=reorder(prin_comp, -varPercent[1:10]), y=varPercent[1:10], )) +
  geom_bar(stat = "identity",fill='steelblue', color="steelblue") + 
  labs(x = "Principal Components", y= "% of Explained Variance") +
  ggtitle("PCA - Scree Plot") +
  theme_minimal()

scores <- as.data.frame(scores[,1:4])
scores$country <- mbov_meta$Country
scores$species <- mbov_meta$Species

ggplot(as.data.frame(scores),aes(x=PC3,y=PC4, fill = country, col = country)) +
  geom_point() +
  theme_minimal()

prab_pca <- prcomp(cloud_prab)
variance <- (prab_pca$sdev)^2
loadings <- prab_pca$rotation
rownames(loadings) <- colnames(cloud_prab)
scores <- prab_pca$x 

varPercent <- variance/sum(variance) * 100 
prin_comp <- paste0("PC",1:10)
data <- data.frame(prin_comp,varPercent[1:10])
ggplot(data, aes(x=reorder(prin_comp, -varPercent[1:10]), y=varPercent[1:10], )) +
  geom_bar(stat = "identity",fill='steelblue', color="steelblue") + 
  labs(x = "Principal Components", y= "% of Explained Variance") +
  ggtitle("PCA - Scree Plot") +
  theme_minimal()

scores <- as.data.frame(scores[,1:4])
scores$country <- mbov_meta$Country
scores$species <- mbov_meta$Species

ggplot(as.data.frame(scores),aes(x=PC3,y=PC4, fill = country, col = country)) +
  geom_point() +
  theme_minimal()
```
So the soft core genes don't give us enough information to make clusters, the shell and cloud genes seem to be better at making this distinction, and they are identical in their output. Maybe we should see what happens if we exlude one (cloud or shell) from the analysis.

```{r}
soft_shell <- subset(accessory_genome,No..isolates > 105 & No..isolates < 693) %>% select(14:(ncol(accessory_genome)))
soft_shell[!(soft_shell=="")] <- 1
soft_shell[soft_shell==""] <- 0
soft_shell <- t(data.matrix(soft_shell))
soft_cloud <- subset(accessory_genome,(No..isolates > 0 & No..isolates < 105) | (No..isolates > 665 & No..isolates < 693)) %>% select(14:(ncol(accessory_genome)))
soft_cloud[!(soft_cloud=="")] <- 1
soft_cloud[soft_cloud==""] <- 0
soft_cloud <- t(data.matrix(soft_cloud))
shell_cloud <- subset(accessory_genome,No..isolates > 0 & No..isolates < 665) %>% select(14:(ncol(accessory_genome)))
shell_cloud[!(shell_cloud=="")] <- 1
shell_cloud[shell_cloud==""] <- 0
shell_cloud <- t(data.matrix(shell_cloud))
```

```{r}
prab_pca <- prcomp(soft_shell)
variance <- (prab_pca$sdev)^2
loadings <- prab_pca$rotation
rownames(loadings) <- colnames(soft_shell)
scores <- prab_pca$x 

varPercent <- variance/sum(variance) * 100 
prin_comp <- paste0("PC",1:10)
data <- data.frame(prin_comp,varPercent[1:10])
ggplot(data, aes(x=reorder(prin_comp, -varPercent[1:10]), y=varPercent[1:10], )) +
  geom_bar(stat = "identity",fill='steelblue', color="steelblue") + 
  labs(x = "Principal Components", y= "% of Explained Variance") +
  ggtitle("PCA - Scree Plot") +
  theme_minimal()

scores <- as.data.frame(scores[,1:4])
scores$country <- mbov_meta$Country
scores$species <- mbov_meta$Species

ggplot(as.data.frame(scores),aes(x=PC3,y=PC4, fill = country, col = country)) +
  geom_point() +
  theme_minimal()

prab_pca <- prcomp(soft_cloud)
variance <- (prab_pca$sdev)^2
loadings <- prab_pca$rotation
rownames(loadings) <- colnames(soft_cloud)
scores <- prab_pca$x 

varPercent <- variance/sum(variance) * 100 
prin_comp <- paste0("PC",1:10)
data <- data.frame(prin_comp,varPercent[1:10])
ggplot(data, aes(x=reorder(prin_comp, -varPercent[1:10]), y=varPercent[1:10], )) +
  geom_bar(stat = "identity",fill='steelblue', color="steelblue") + 
  labs(x = "Principal Components", y= "% of Explained Variance") +
  ggtitle("PCA - Scree Plot") +
  theme_minimal()

scores <- as.data.frame(scores[,1:4])
scores$country <- mbov_meta$Country
scores$species <- mbov_meta$Species

ggplot(as.data.frame(scores),aes(x=PC3,y=PC4, fill = country, col = country)) +
  geom_point() +
  theme_minimal()

prab_pca <- prcomp(shell_cloud)
variance <- (prab_pca$sdev)^2
loadings <- prab_pca$rotation
rownames(loadings) <- colnames(shell_cloud)
scores <- prab_pca$x 

varPercent <- variance/sum(variance) * 100 
prin_comp <- paste0("PC",1:10)
data <- data.frame(prin_comp,varPercent[1:10])
ggplot(data, aes(x=reorder(prin_comp, -varPercent[1:10]), y=varPercent[1:10], )) +
  geom_bar(stat = "identity",fill='steelblue', color="steelblue") + 
  labs(x = "Principal Components", y= "% of Explained Variance") +
  ggtitle("PCA - Scree Plot") +
  theme_minimal()

scores <- as.data.frame(scores[,1:4])
scores$country <- mbov_meta$Country
scores$species <- mbov_meta$Species

ggplot(as.data.frame(scores),aes(x=PC3,y=PC4, fill = country, col = country)) +
  geom_point() +
  theme_minimal()
```
These plots are intriguing because the clustering we see with all the is closest when we include the shell genes! so for this type of analysis, shell genes provides the most information in order to cluster using PCA. My interpretation: The shell genes carry the most variation in presence/absence, so are the most useful for the analysis. Next, the cloud genes provide the necessary information to improve this clustering (the unique genes.)


```{r}
loadings <- prab_pca$rotation
loadings <- data.frame(round(abs(loadings[,3:4]), 2))

ggplot(loadings,aes(x=PC3,y=PC4)) +
  geom_point() +
  geom_jitter() +
  theme_minimal()

#I'm searching for ways to conduct loading analysis, BUT FOR NOW, I will just get the higher values. To me, I will get (over 0.08, over 0.0625)

filtered_loadings <- subset(loadings, PC3 > 0.02 & loadings$PC4 > 0.02)

ggplot(filtered_loadings,aes(x=PC3,y=PC4)) +
  geom_point() +
  geom_jitter() +
  theme_minimal()

key_COGs <- data.frame(filtered_loadings,stringsAsFactors = FALSE)
key_COGs$COG <- rownames(key_COGs)
```

```{r}
library(dplyr)

#now, let's see how these COGs relate to genome annotations
auxil$COG <- rownames(auxil)
key_COGS <- key_COGs %>% left_join(auxil,by = "COG")
key_COGS$Annotation
```

